Response Patterns

### 2025-01-09 misc/indiv.qmd
todays  updates - intro, discussion, few methods tweaks

s1_agg <- s1 |> 
  filter(appliance !="Total kWh") |> 
  group_by(id,refClass,state,block,plan,calc,edu,pct_goal) |> 
  summarise(total_kWh = sum(value),orig_kWh=sum(family), 
            pct_change = round((orig_kWh-total_kWh)/orig_kWh,3), 
            n_change = sum(value!=family),
            state_p_dif=mean(state_p_dif),
            state_f_dif=mean(state_f_dif),
            n_less_avg = sum(less_avg),
            duration=first(Duration__in_seconds_)) |> 
  mutate(matched_goal = (pct_change == pct_goal), 
             error = pct_change - pct_goal,
            abs_error = abs(error))
```

>>>>

s1_agg <- s1 |> 
    filter(appliance !="Total kWh") |> 
    group_by(id,refClass,state,block,plan,edu,pct_goal,calc) |> 
    summarise(total_kWh = sum(value),orig_kWh=sum(family), 
                pct_change = abs(round((orig_kWh-total_kWh)/orig_kWh,3)), 
                n_change = sum(value!=family),
                state_p_dif=mean(state_p_dif),
                state_f_dif=mean(state_f_dif),
                n_less_avg = sum(less_avg),
                duration=first(Duration__in_seconds_)) |> 
    mutate(matched_goal = (pct_change == pct_goal), 
                error = pct_change - pct_goal,
                abs_error = abs(error),
                log_abs_error=log(abs(error)+.007), 
                close_match = abs_error <= 0.02) |>
    ungroup() |> # Add ungroup here
        mutate(
            accuracy_level = factor(
                case_when(
                    abs_error == 0.00 ~ "Exact match",
                    abs_error <= 0.05 ~ "0.01-5% error",
                    TRUE ~ "Over 5% error"  # Capture all remaining cases
                ), 
                levels = c("Exact match", "0.01-5% error", "Over 5% error"),
                ordered = TRUE
            )
        ) |> relocate(accuracy_level, .after= "pct_change")

s1_agg4 <- s1_agg |> group_by(id,refClass,calc) |> 
    mutate(n_accuracy = n_distinct(accuracy_level)) |> 
    summarise(
    mg=sum(matched_goal),
    mgc=sum(close_match),
    n=n(), 
    pct=mg/n,
    pct_close=mgc/n,
    mean_pct_change=mean(pct_change),
    mean_abs_error=mean(abs_error),
    mean_log_abs_error=mean(log_abs_error),
    n_accuracy=first(n_accuracy)) |> 
    mutate(accuracy_level = factor(
            case_when(
                mean_abs_error < 0.02 ~ "Exact match",
                mean_abs_error <= 0.05 ~ "0.01-5% error",
                TRUE ~ "Over 5% error"  # Capture all remaining cases
            ), 
            levels = c("Exact match", ".01-5% error", "Over 5% error"),
            ordered = TRUE
        ))
```

# Clustering

# A tibble: 4,560 × 17
# Groups:   id, block [380]
   id     block refClass appliance pct_goal rounded state    calc       edu                         els   pct family state_avg plan  value change state_dif
   <chr>  <dbl> <fct>    <chr>     <chr>    <chr>   <chr>    <chr>      <fct>                     <dbl> <dbl>  <dbl>     <dbl> <fct> <dbl>  <dbl>     <dbl>
 1 s2_100     1 kWh      AC        10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1   1000       498 plan1   500   -500         2
 2 s2_100     1 kWh      AC        10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1   1000       498 plan2   300   -700      -198
 3 s2_100     1 kWh      HEAT      10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1  18000     16411 plan1 15000  -3000     -1411
 4 s2_100     1 kWh      HEAT      10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1  18000     16411 plan2 12000  -6000     -4411
 5 s2_100     1 kWh      WATER     10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1  12000      5832 plan1  6000  -6000       168
 6 s2_100     1 kWh      WATER     10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1  12000      5832 plan2  2000 -10000     -3832
 7 s2_100     1 kWh      REFRIG    10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1   1000      1142 plan1  2000   1000       858
 8 s2_100     1 kWh      REFRIG    10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1   1000      1142 plan2   500   -500      -642

```{r}
#| eval: false
# widen by appliance, with separate row for value, family, and stage_avg

s2 <- s2_long #|> select(-family,-state_avg,-edu,-change,-state_dif) |> pivot_wider(names_from = appliance, values_from = c(value)) 
s2s <- s2_long |> select(-family,-state_avg,-edu,-change,-value) |> pivot_wider(names_from = appliance, values_from = c(state_dif)) |> select(id,block,refClass,pct,plan,AC:OTHER)
# s2s
#    id     block refClass     pct plan     AC   HEAT WATER REFRIG OTHER
#    <chr>  <dbl> <fct>      <dbl> <fct> <dbl>  <dbl> <dbl>  <dbl> <dbl>
#  1 s2_100     1 kWh         0.1  plan1     2  -1411   168    858   348
#  2 s2_100     1 kWh         0.1  plan2  -198  -4411 -3832   -642 -6152
#  3 s2_100     2 kWh         0.15 plan1  1324     19  -139    320  -424
#  4 s2_100     2 kWh         0.15 plan2   324   -981 -1139   -318 -1424
#  5 s2_101     1 kWh         0.1  plan1   102   -411  2168   -142   398
#  6 s2_101     1 kWh         0.1  plan2   502  -1411   168   -142  1348
#  7 s2_101     2 kWh         0.15 plan1   751    901  -396    682 -1883
#  8 s2_101     2 kWh         0.15 plan2  2251    401   104    682 -1383
#  9 s2_102     1 Percentage  0.1  plan1   958 -12954 -5354    200 -5417
# 10 s2_102     1 Percentage  0.1  plan2   890 -15435 -4956   8544  3225

# clustering planning strategies - kmeans

s2s |>  ungroup() |> select(-id,-block,-refClass,-pct,-plan) |> scale(center = TRUE, scale = TRUE) |> kmeans(centers = 3) 
s2s_clust <- s2s |>  ungroup() |> select(-id,-block,-refClass,-pct,-plan) |> scale(center = TRUE, scale = TRUE) |> kmeans(centers = 3) %>% .$cluster |> as.factor() 
s2s <- cbind(s2s,c=s2s_clust)

s2s |>    group_by(c) |>
  summarise(ac=mean(AC),heat=mean(HEAT),water=mean(WATER),refrig=mean(REFRIG),other=mean(OTHER)) |>
  pivot_longer(cols = c(ac:other), names_to = "Appliance", values_to = "Mean") |>
  ggplot(aes(x=Appliance,y=Mean,col=c)) + 
  geom_point() + 
  facet_wrap(~c) +
  theme_minimal()

s2s <- s2_long |> select(-family,-state_avg,-edu,-change,-state_dif) |> pivot_wider(names_from = appliance, values_from = c(value))  |> select(id,block,refClass,pct,plan,AC:OTHER)

s2s |>  ungroup() |> select(-id,-block,-refClass,-pct,-plan) |> scale(center = TRUE, scale = TRUE) |> kmeans(centers = 3)

s2s_clust <- s2s |>  ungroup() |> select(-id,-block,-refClass,-pct,-plan) |> scale(center = TRUE, scale = TRUE) |> kmeans(centers = 3) %>% .$cluster |> as.factor()

s2s <- cbind(s2s,c=s2s_clust) |> left_join(s2, join_by(id,refClass,block,plan,pct))
s2s |>  group_by(c) |>
  summarise(ac=mean(AC),heat=mean(HEAT),water=mean(WATER),refrig=mean(REFRIG),other=mean(OTHER)) |>
  pivot_longer(cols = c(ac:other), names_to = "Appliance", values_to = "Mean") |>
  ggplot(aes(x=Appliance,y=Mean,col=c)) + 
  geom_point() + 
  geom_point(aes(y=state_avg),color="black") +
  facet_wrap(~c) +
  theme_minimal()

s2 |> group_by(appliance,family,state,state_avg) |> select(appliance,family_prior_yr=family,state,state_avg)
```
